---
title: "Getting started with daddyR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with daddyR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

```{r setup}
library(daddyR)
```

## The scientific question

Given a scalar time series `x(t)`, daddyR estimates the drift and diffusion
functions of the stochastic differential equation

$$ \frac{dx}{dt} = f(x) + g(x)\,\eta(t) $$

that best describes its dynamics, where $\eta(t)$ is Gaussian white noise.
$f$ (the *drift*) governs the deterministic part of the dynamics; $g^2$ (the
*diffusion*) governs the strength of the stochastic part. This is the method
of Nabeel et al. (2025), *The American Naturalist* 205:E100-E117.

## A worked example

We'll use one of the bundled example datasets: a simulated time series from
a simple pairwise-interaction model (the kind used throughout the original
paper's synthetic examples).

```{r}
d <- dd_load_sample_data("scalar-pairwise")
str(d)
```

`dd_analyse()` does the core estimation: binning the data by its own value
and computing the average drift and diffusion within each bin.

```{r}
dd <- dd_analyse(d$x, t = d$t)
dd
```

`dd$Dt` was chosen automatically from the series' autocorrelation time,
since we didn't specify one (see `?dd_analyse` and `DESIGN_DECISIONS.md` for
why this differs from PyDaddy's own default). The binned estimates
themselves are ordinary data frames:

```{r}
head(dd_drift(dd))
head(dd_diffusion(dd))
```

## Fitting analytical expressions

`dd_fit()` fits a sparse polynomial to the drift or diffusion function,
using sequentially-thresholded ridge regression (the same approach as
PyDaddy, minus the standard-error-computation bug documented in
`DESIGN_DECISIONS.md`). It returns an updated copy of the `daddy` object
with the fit attached — R objects aren't mutated in place, so you need to
reassign:

```{r}
dd <- dd_fit(dd, "drift", degree = 3, threshold = 0.01)
dd <- dd_fit(dd, "diffusion", degree = 2, threshold = 0.01)
dd
```

## Visualizing the result

```{r, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE)}
plot(dd)
```

## Diagnostics

Is the extracted noise consistent with a Gaussian white-noise assumption?

```{r}
dd_gaussianity_test(dd$estimate$diff_series)
```

And the autocorrelation structure of the raw series:

```{r}
ac <- dd_autocorrelation(d$x)
ac$autocorr_time
```

## Exporting results

```{r, eval = FALSE}
dd_export_data(dd, file = "drift_diffusion.csv")
```
